Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2021-12-02 Manuscript/Manuscript.tex

@@

\begin{center}\includegraphics{Manuscript_files/figure-latex/patterns-1} \end{center}

\end{CodeChunk}

@@

\hypertarget{the-incomplete-data}{%
\subsubsection{The incomplete data}\label{the-incomplete-data}}

\begin{itemize}
\tightlist
\item
  What does the missingness look like? Induced MAR/MNAR missingness.
  Missing data pattern:
\end{itemize}

@@

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_pat-1} \end{center}

\end{CodeChunk}

\begin{itemize}
\item
  What are the ICCs? For \texttt{popular} the ICC is 0.33. For
  \texttt{popteach} it is 0.31. It would be wise to use multilevel
  modeling.
\item
  Does the missing data of \texttt{popular} depend on \texttt{popteach}?
  Does the missingness in teacher popularity depend on pupil popularity?
  -\textgreater{} Check this by making a histogram of \texttt{popteach}
  separately for the pupils with known popularity and missing
  popularity, and the other way around.
\end{itemize}

@@

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_dist-1} \end{center}

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_dist-2} \end{center}

\end{CodeChunk}

\begin{itemize}
\tightlist
\item
  We do see that the distribution for the missing \texttt{popular} is
  further to the right than the distribution for observed
  \texttt{popular}. This would indicate a right-tailed MAR missingness.
  In fact this is exactly what happens, because we created the
  missingness in these data ourselves. But we made it observable by
  examining the relations between the missingness in popular and the
  observed data in \texttt{popteach}. There is also a dependency between
  the missingness in teacher popularity and pupil popularity. The
  relation seems to be right-tailed as well.
\end{itemize}

\hypertarget{the-imputation-model}{%
\subsubsection{The imputation model}\label{the-imputation-model}}

\begin{itemize}
\item
  We could ignore the multilevel structure of the data. This is likely
  invalid, given the high ICCs, but we'll do it anyways.
\item
  We'll use predictive mean matching to impute the continuous variables
  (some appear to be somewhat ordinal), and logistic regression to
  impute the binary variable \texttt{sex}. We do not use the observation
  identifier \texttt{pupil} or cluster identifier \texttt{class} as
  predictors to impute other variables.
\end{itemize}

\begin{CodeChunk}
\begin{CodeInput}
R> # dry run to get imputation parameters
R> ini <- mice(pop, maxit = 0)
R> 
R> # extract predictor matrix and adjust
R> pred <- ini$pred
R> pred[, c("class", "pupil")] <- 0
R> pred 
\end{CodeInput}
\begin{CodeOutput}
         pupil class extrav sex texp popular popteach school
pupil        0     0      1   1    1       1        1      1
class        0     0      1   1    1       1        1      1
extrav       0     0      0   1    1       1        1      1
sex          0     0      1   0    1       1        1      1
texp         0     0      1   1    0       1        1      1
popular      0     0      1   1    1       0        1      1
popteach     0     0      1   1    1       1        0      1
school       0     0      1   1    1       1        1      0
\end{CodeOutput}
\begin{CodeInput}
R> # impute the data, ignoring the cluster structure
R> imp_ignored <- mice(pop, maxit = 10, pred = pred, print = FALSE)
R> 
R> # check convergence of the imputation model
R> plot(imp_ignored)
\end{CodeInput}

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-ignored-1} \end{center}

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-ignored-2} \end{center}

\begin{CodeInput}
R> # compare descriptives before and after imputation
R> psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65   11.0   1  26.0  5.97
class*   2000 50.37   51.0   1 100.0 29.08
extrav   1484  5.31    5.0   1  10.0  1.29
sex*     1504  1.56    2.0   1   2.0  0.50
texp     1024 11.80   12.0   2  25.0  6.26
popular  1490  4.83    4.8   0   9.1  1.34
popteach 1472  4.83    5.0   1  10.0  1.36
school   2000  5.54    6.0   1  10.0  2.89
\end{CodeOutput}
\begin{CodeInput}
R> psych::describe(mice::complete(imp_ignored))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65     11   1  26.0  5.97
class*   2000 50.37     51   1 100.0 29.08
extrav   2000  5.23      5   1  10.0  1.25
sex*     2000  1.51      2   1   2.0  0.50
texp     2000 12.55     12   2  25.0  6.27
popular  2000  4.99      5   0   9.1  1.32
popteach 2000  5.01      5   1  10.0  1.38
school   2000  5.54      6   1  10.0  2.89
\end{CodeOutput}
\begin{CodeInput}
R> # TODO: add stripplot with boxplot overlay instead of the tables (make pooled one thick on top)
R> # TODO: pool mean median and sd
R> # feedback Stef: numbers, continuous statistics such as means, and uncertainty estimates. So we can pool the sd's. And leave out the min and max, because those are not normally distr.
R> # TODO: add FMI for each of the estimates? at least for the mean
R> 
R> # further inspection of the imputations
R> densityplot(imp_ignored)
\end{CodeInput}

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-ignored-3} \end{center}

\begin{CodeInput}
R> # compare ICCs before and after imputation
R> ICCs <- data.frame(
+   vars = c("popular", "popteach", "texp"), 
+   incomplete = c(multilevel::ICC1(aov(popular ~ class, pop)), 
+                multilevel::ICC1(aov(popteach ~ class, pop)),
+                multilevel::ICC1(aov(texp ~ class, pop))), 
+   ignored = c(multilevel::ICC1(aov(popular ~ class, complete(imp_ignored))), 
+               multilevel::ICC1(aov(popteach ~ class, complete(imp_ignored))), 
+               multilevel::ICC1(aov(texp ~ class, complete(imp_ignored))))
+   )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored
1  popular  0.3280070 0.2754195
2 popteach  0.3138658 0.2467415
3     texp  1.0000000 0.4443760
\end{CodeOutput}
\end{CodeChunk}

\begin{itemize}
\item
  As the original ICCs show, 100\% of the variance in \texttt{texp} can
  be attributed to the clustering variable \texttt{class}. This tells us
  that the multilevel structure of the data should be taken into
  account. If we don't, we'll end up with incorrect imputations, biasing
  the effect of the clusters towards zero.
\item
  We can also observe that the teacher experience increases slightly
  after imputation. This is due to the MNAR missingness in
  \texttt{texp}. Higher values for \texttt{texp} have a larger
  probability to be missing. This may not a problem, however, if at
  least one pupil in each class has teacher experience recorded, we can
  deductively impute the correct (i.e.~true) value for every pupil in
  the class.
\item
  We'll now use \texttt{class} as a predictor to impute all other
  variables.
\end{itemize}

\begin{CodeChunk}
\begin{CodeInput}
R> # adjust the predictor matrix
R> pred <- ini$pred 
R> pred[, "pupil"] <- 0
R> pred
\end{CodeInput}
\begin{CodeOutput}
         pupil class extrav sex texp popular popteach school
pupil        0     1      1   1    1       1        1      1
class        0     0      1   1    1       1        1      1
extrav       0     1      0   1    1       1        1      1
sex          0     1      1   0    1       1        1      1
texp         0     1      1   1    0       1        1      1
popular      0     1      1   1    1       0        1      1
popteach     0     1      1   1    1       1        0      1
school       0     1      1   1    1       1        1      0
\end{CodeOutput}
\begin{CodeInput}
R> # impute the data, cluster as predictor
R> imp_predictor <- mice(pop, maxit = 10, pred = pred, print = FALSE)
\end{CodeInput}
\begin{CodeOutput}
Warning: Number of logged events: 312
\end{CodeOutput}
\begin{CodeInput}
R> # check logged events
R> head(imp_predictor$loggedEvents)
\end{CodeInput}
\begin{CodeOutput}
  it im      dep   meth
1  1  1   extrav    pmm
2  1  1      sex logreg
3  1  1     texp    pmm
4  1  1  popular    pmm
5  1  1  popular    pmm
6  1  1 popteach    pmm
                                                                                                                                                                                                                                                       out
1                                                                                                                                                                                                                                                   school
2                                                                                                                                                                                                                                                   school
3                                                                                                                                                                                                                                                   school
4                                                                                                                                                                                                                                                     texp
5 mice detected that your data are (nearly) multi-collinear.\nIt applied a ridge penalty to continue calculations, but the results can be unstable.\nDoes your dataset contain duplicates, linear transformation, or factors with unique respondent names?
6                                                                                                                                                                                                                                             texp, school
\end{CodeOutput}
\begin{CodeInput}
R> ## "The mice() function detects multicollinearity, and solves the problem by removing one or more predictors for the model", in this case texp is removed as predictor of popular and popteach.
R> 
R> # check convergence of the imputation model
R> plot(imp_predictor)
\end{CodeInput}

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-predictor-1} \end{center}

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-predictor-2} \end{center}

\begin{CodeInput}
R> # compare descriptives before and after imputation
R> psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65   11.0   1  26.0  5.97
class*   2000 50.37   51.0   1 100.0 29.08
extrav   1484  5.31    5.0   1  10.0  1.29
sex*     1504  1.56    2.0   1   2.0  0.50
texp     1024 11.80   12.0   2  25.0  6.26
popular  1490  4.83    4.8   0   9.1  1.34
popteach 1472  4.83    5.0   1  10.0  1.36
school   2000  5.54    6.0   1  10.0  2.89
\end{CodeOutput}
\begin{CodeInput}
R> psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65   11.0   1  26.0  5.97
class*   2000 50.37   51.0   1 100.0 29.08
extrav   2000  5.22    5.0   1  10.0  1.29
sex*     2000  1.53    2.0   1   2.0  0.50
texp     2000 14.26   15.0   2  25.0  6.55
popular  2000  5.59    5.4   0   9.1  1.81
popteach 2000  5.03    5.0   1  10.0  1.40
school   2000  5.54    6.0   1  10.0  2.89
\end{CodeOutput}
\begin{CodeInput}
R> # further inspection of the imputations
R> densityplot(imp_predictor)
\end{CodeInput}

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-predictor-3} \end{center}

\begin{CodeInput}
R> # compare ICCs before and after imputation
R> ICCs <- ICCs %>% cbind(
+            predictor = c(multilevel::ICC1(aov(popular ~ class, complete(imp_predictor))), 
+                         multilevel::ICC1(aov(popteach ~ class, complete(imp_predictor))), 
+                         multilevel::ICC1(aov(texp ~ class, complete(imp_predictor))))
+            )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored predictor
1  popular  0.3280070 0.2754195 0.2496278
2 popteach  0.3138658 0.2467415 0.3483987
3     texp  1.0000000 0.4443760 1.0000000
\end{CodeOutput}
\end{CodeChunk}

\begin{itemize}
\item
  Now, we can clearly see that the imputed values of \texttt{texp} are
  higher than the observed values, which is in line with right-tailed
  MNAR.
\item
  The ICCs are way more in line with the ICCs in the incomplete data.
  But this is a quick and dirty way of imputing multilevel data. We
  \emph{should} be using a multilevel model.
\end{itemize}

@@

\begin{CodeChunk}
\begin{CodeOutput}
 /\     /\
{  `---'  }
{  O   O  }
==>  V <==  No need for mice. This data set is completely observed.
 \  \|/  /
  `-----'
\end{CodeOutput}

\begin{center}\includegraphics{Manuscript_files/figure-latex/impact-1} \end{center}

\end{CodeChunk}

>>>>

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{Manuscript_files/figure-latex/patterns-1}

}

\caption[Missingness in multilevel data]{Missingness in multilevel data}\label{fig:patterns}
\end{figure}
\end{CodeChunk}

@@

@@

\hypertarget{incomplete-data}{%
\subsubsection{Incomplete data}\label{incomplete-data}}

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{Manuscript_files/figure-latex/pop_pat-1}

}

\caption[Missing data pattern in the popularity data]{Missing data pattern in the popularity data}\label{fig:pop_pat}
\end{figure}
\end{CodeChunk}

The popularity data is created such that there are strong relations
between the incomplete variables and the clustering variable
\texttt{class}. We can express this using the intra-class correlation
(ICC). For \texttt{popular} the ICC is 0.33. For \texttt{popteach} it is
0.31. It would thus be wise to use multilevel modeling.

The missingness in this dataset is induced conform MAR and MNAR
mechanisms. The missing data pattern, Figure \ref{fig:pop_pat}, shows
the systematic nature of the missingness.

To develop the best imputation model, we need to know whether the
missingness in one variable depends on the observed values of other
variables. Visual inspection usually suffices. We'll highlight only two
variables to illustrate, but ideally one would inspect all relations.
The questions we'll ask are: `Does the missing data of \texttt{popular}
depend on \texttt{popteach}?' and `Does the missingness in teacher
popularity depend on pupil popularity?' We'll evaluate this by making a
histogram of \texttt{popteach} separately for the pupils with known
popularity and missing popularity, and the other way around.

In Figure \ref{fig:pop_dist} we see that the distribution for the
missing \texttt{popular} is further to the right than the distribution
for observed \texttt{popular}. This would indicate a right-tailed MAR
missingness. In fact, this is exactly what happens, because the
missingness in these data was created manually. Now, we've made it
observable by examining the relations between the missingness in popular
and the observed data in \texttt{popteach}. There is also a dependency
between the missingness in teacher popularity and pupil popularity. The
relation seems to be right-tailed as well.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{Manuscript_files/figure-latex/pop_dist-1}

}

\caption[Conditional distributions in the popularity data]{Conditional distributions in the popularity data}\label{fig:pop_dist-1}
\end{figure}
\begin{figure}

{\centering \includegraphics{Manuscript_files/figure-latex/pop_dist-2}

}

\caption[Conditional distributions in the popularity data]{Conditional distributions in the popularity data}\label{fig:pop_dist-2}
\end{figure}
\end{CodeChunk}

\hypertarget{imputation-model}{%
\subsubsection{Imputation model}\label{imputation-model}}

The first imputation model that we'll use is likely to be invalid. In
this model, we ignore the multilevel structure of the data, despite the
high ICCs. This is purely to illustrate the effects of ignoring the
clustering in our imputation effort.

We'll use predictive mean matching to impute the continuous variables
and logistic regression to impute the binary variable \texttt{sex}. We
do not use the observation identifier \texttt{pupil} or cluster
identifier \texttt{class} as predictors to impute other variables.

\begin{CodeChunk}
\begin{CodeInput}
R> # dry run to get imputation parameters
R> ini <- mice(pop, maxit = 0)
R> 
R> # extract predictor matrix and adjust
R> pred <- ini$pred
R> pred[, c("class", "pupil")] <- 0
R> 
R> # impute the data, ignoring the cluster structure
R> imp_ignored <- mice(pop, maxit = 10, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

\hypertarget{imputed-data}{%
\subsubsection{Imputed data}\label{imputed-data}}

@@

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_ignored_eval-1} \end{center}

\begin{CodeOutput}
      vars incomplete   ignored
1  popular  0.3280070 0.2958571
2 popteach  0.3138658 0.2561198
3     texp  1.0000000 0.4696024
\end{CodeOutput}
\end{CodeChunk}

As the original ICCs show, 100\% of the variance in \texttt{texp} can be
attributed to the clustering variable \texttt{class}. This tells us that
the multilevel structure of the data should be taken into account. If we
don't, we'll end up with incorrect imputations, biasing the effect of
the clusters towards zero.

We can also observe that the teacher experience increases slightly after
imputation. This is due to the MNAR missingness in \texttt{texp}. Higher
values for \texttt{texp} have a larger probability to be missing. This
may not a problem, however, if at least one pupil in each class has
teacher experience recorded, we can deductively impute the correct
(i.e.~true) value for every pupil in the class.

\hypertarget{imputation-model-1}{%
\subsubsection{Imputation model}\label{imputation-model-1}}

We'll now use \texttt{class} as a predictor to impute all other
variables. This is still not recommended practice, since it only works
under certain circumstances and results may be biased. But at least, it
includes some multilevel aspect. Colloquially, this is `multilevel
imputation for dummies'.

\hypertarget{imputed-data-1}{%
\subsubsection{Imputed data}\label{imputed-data-1}}

@@

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_predictor_eval-1} \end{center}

\begin{CodeOutput}
      vars incomplete   ignored predictor
1  popular  0.3280070 0.2958571 0.3581435
2 popteach  0.3138658 0.2561198 0.3446899
3     texp  1.0000000 0.4696024 1.0000000
\end{CodeOutput}
\end{CodeChunk}

Now, we can clearly see that the imputed values of \texttt{texp} are
higher than the observed values, which is in line with right-tailed
MNAR.

The ICCs are way more in line with the ICCs in the incomplete data. But
this is a quick and dirty way of imputing multilevel data. We
\emph{should} be using a multilevel model.

\hypertarget{imputation-model-2}{%
\subsubsection{Imputation model}\label{imputation-model-2}}

To include\ldots{}

@@
